---
title: "Re-analysing Patchett 2020 data"
author: "Anne-Lise Gérard"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source codes: https://github.com/Anne-LiseGerard/dftd_rnaseq
Data: https://doi.org/10.1007/s00018-019-03259-2

```{r,pkg}

suppressPackageStartupMessages({
    library("reshape2")
    library("gplots")
    library("DESeq2")
    library("mitch")
    library("limma")
    library("kableExtra")
    library("dplyr")
    library("tidyr")
    library("ComplexHeatmap")
    library("RColorBrewer")
    #library("circlize")
    #library("pathview")
    library("stringr")
    #library("grid")
})

```

## Background

Goal: re-analyse previously published DFTD tumour biopsy transcriptomes and compare to DFTD cell line transcriptomes, focusing on genes involved in metabolism. Below is a list of the samples for this project.

```{r, ss}

ss <- read.table("../ss_patchett.txt",sep="\t",fill=TRUE,header=TRUE)
ss$DFT <- as.factor(ss$DFT)

ss %>%
  kbl(caption="Sample sheet for all samples") %>%
  kable_paper("hover", full_width = F)

```

## Functions

```{r, func}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange,
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=1, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=1, col="red")
  mtext(SUBHEADER,cex = 0.7)
  grid()
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=1,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval")
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=1,pch=19,col="red")
}

```

## Load data

Import data from the aligner. 

```{r, import}

tmp <- read.table("../fastq/3col_patchett.tsv.gz")
tmp$V1 <- gsub('.fastq-trimmed.fastq','',tmp$V1)
x <- as.data.frame(acast(tmp, V2~V1, value.var="V3",fun.aggregate=sum))

tmp_old <- read.table("../fastq/3col_old.tsv.gz")
tmp_old$V1 <- gsub('.fastq-trimmed.fastq','',tmp_old$V1)
x_old <- as.data.frame(acast(tmp_old, V2~V1, value.var="V3",fun.aggregate=sum))

```

Load gene names.

```{r, genenames}

gn <- read.table("../ref/Sarcophilus_harrisii.mSarHar1.11.cdna+ncrna.genenames.tsv",fill=TRUE)
gn <- gn[order(gn$V1),]

gn_old <- read.table("../ref/Sarcophilus_harrisii.DEVIL7.0.cdna+ncrna.genenames.tsv",fill=TRUE)
gn_old <- gn_old[order(gn_old$V1),]

```

```{r, homology}

hm <- read.table("../ref/mart_export_ensembl109_2023-07-14.txt",sep="\t",header=TRUE)

hm_old <- read.table("../ref/mart_export_ensembl101_2023-09-06.txt",sep="\t",header=TRUE)

```

Now need to collapse transcript data to genes.

```{r, collapse}

x$gene <- paste(gn$V2,gn$V3)

y <- aggregate(. ~ gene,x,sum)

rownames(y) <- y$gene
y$gene = NULL


x_old$gene <- paste(gn_old$V2,gn_old$V3)

y_old <- aggregate(. ~ gene,x_old,sum)

rownames(y_old) <- y_old$gene
y_old$gene = NULL

```

## Quality control

Samples with <1M reads should be omitted.
Will also round values to integers.

```{r, qc1}

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))
barplot(cs,main="All samples",horiz=TRUE,las=1)
abline(v=1e7,col="red",lty=2)

y <- round(y)


cs_old <- colSums(y_old)
cs_old <- cs_old[order(cs_old)]

par(mar=c(5,10,5,2))
barplot(cs_old,main="All samples - old genome ",horiz=TRUE,las=1, xli=c(0,4e+07))
abline(v=1e7,col="red",lty=2)

y_old <- round(y_old)

```

## MDS

This will help us to visualise the sources of variability in the overall dataset.

Plot MDS. Also fix the sample names.

Plot MDS by DFT.

```{r, plotmds1}

par(mar = c(5.1, 4.1, 4.1, 2.1))

colnames(y) <- ss$sample_id

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))
barplot(cs,main="All samples",horiz=TRUE,las=1)


par(mar = c(5.1, 4.1, 4.1, 2.1))

colnames(y_old) <- ss$sample_id[1:8]

cs_old <- colSums(y_old)
cs_old <- cs_old[order(cs_old)]

par(mar=c(5,10,5,2))
barplot(cs_old,main="All samples - old genome",horiz=TRUE,las=1, xlim=c(0,4e+07))


par(mar = c(5.1, 4.1, 4.1, 2.1) )

cols <- ss$DFT
cols <- gsub("DFT1","lightblue",cols)
cols <- gsub("DFT2","pink",cols)
mymds <- plotMDS(y,plot=FALSE)

mymds_old <- plotMDS(y_old,plot=FALSE)

# fix the xlims
XMIN=min(mymds$x)
XMAX=max(mymds$x)
XMID=(XMAX+XMIN)/2
XMIN <- XMID + (XMIN-XMID)*1.1
XMAX <- XMID+(XMAX-XMID)*1.1
plotMDS(mymds,pch=19,cex=3,col=cols,main="MDS plot",xlim=c(XMIN,XMAX))
text(mymds,labels=colnames(y))
mtext("blue=DFT1, pink=DFT2")

plotMDS(mymds_old,pch=19,cex=3,col=cols,main="MDS plot - old genome",xlim=c(XMIN,XMAX))
text(mymds_old,labels=colnames(y_old))
mtext("blue=DFT1, pink=DFT2")

y_DFT1 <- y[,colnames(y) %in% c("sha_DFT1_1-1","sha_DFT1_1-2","sha_DFT1_2-1","sha_DFT1_2-2",
                                "C5065_UT_01", "C5065_UT_02")]
y_DFT2 <- y[,colnames(y) %in% c("sha_DFT2_1-1","sha_DFT2_1-2","sha_DFT2_2-1","sha_DFT2_2-2",
                                "sha_DFT2_RV_1-1","sha_DFT2_RV_1-2","sha_DFT2_RV_2-1","sha_DFT2_RV_2-2")] 

y_biopsy <- y[,colnames(y) %in% c("sha_DFT1_1-1","sha_DFT1_1-2","sha_DFT1_2-1","sha_DFT1_2-2",
                                  "sha_DFT2_1-1","sha_DFT2_1-2","sha_DFT2_2-1","sha_DFT2_2-2")]
y_cline <- y[,colnames(y) %in% c( "C5065_UT_01", "C5065_UT_02",
                                "sha_DFT2_RV_1-1","sha_DFT2_RV_1-2","sha_DFT2_RV_2-1","sha_DFT2_RV_2-2")] 

mdsDFT1 <- plotMDS(y_DFT1,plot=FALSE)
plotMDS(mdsDFT1,pch=19,cex=3,col="pink",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT1,labels=colnames(y_DFT1))
mtext("DFT1")

mdsDFT2 <- plotMDS(y_DFT2,plot=FALSE)
plotMDS(mdsDFT2,pch=19,cex=3,col="lightblue",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT2, labels=colnames(y_DFT2))
mtext("DFT2")

mds_biopsy <- plotMDS(y_biopsy,plot=FALSE)
plotMDS(mds_biopsy,pch=19,cex=3,col=cols,main="MDS plot",xlim=c(XMIN,XMAX))
text(mds_biopsy,labels=colnames(y_biopsy))
mtext("Tumour biopsies: blue=DFT1, pink=DFT2")

mds_cline <- plotMDS(y_cline,plot=FALSE)
plotMDS(mds_cline,pch=19,cex=3,col=cols,main="MDS plot",xlim=c(XMIN,XMAX))
text(mds_cline,labels=colnames(y_cline))
mtext("Cell lines: blue=DFT1, pink=DFT2")

```

## DESeq2

Sum replicates then run differential expression analysis.
For now, ignore alignment on old genome and cell line data.

```{r, de}

y_cell <- y[,9:14]
y <- y[,1:8]

# combine replicates for each sample
DFT1_1 <- rowSums(y[,ss$sample=="DFT1_1"])
DFT1_2 <- rowSums(y[,ss$sample=="DFT1_2"])
DFT2_1 <- rowSums(y[,ss$sample=="DFT2_1"])
DFT2_2 <- rowSums(y[,ss$sample=="DFT2_2"])

y2 <- data.frame(DFT1_1,DFT1_2,DFT2_1,DFT2_2)
ss2 <- as.data.frame(colnames(y2))
colnames(ss2) <- "sample"
ss2$DFT <- factor(c("DFT1","DFT1","DFT2","DFT2"))

y2 <- y2[which(rowMeans(y2)>10),] # remove genes with average count < 10
dim(y2)

dds <- DESeqDataSetFromMatrix(countData = y2 , colData = ss2, design = ~ DFT )
dds2 <- dds
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge2 <- dge
sig <- subset(dge,padj<0.05)
sig2_up <- rownames(subset(sig,log2FoldChange>0))
sig2_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig2_up)
length(sig2_dn)

```

```{r,deviz}

maplot(dge2,"DFT1 vs DFT2 biopsies")

make_volcano(dge2,"DFT1 vs DFT2 biopsies")

sig[1:50,1:6] %>%
  kbl(caption="DFT1 vs DFT2 biopsies") %>%
  kable_paper("hover", full_width = F)

write.table(dge2,file="dge2.tsv",sep="\t")

# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols[3:6],mar=c(8,14),
  cexRow=0.7,cexCol=1)

```
## Enrichment analysis

Need to get human homologs of these genes.
These were obtained from Ensembl v109 biomart (website).

```{r}

rownames(dge2) <- sapply(strsplit(rownames(dge2),"\\."),"[[",1)

hm2 <- hm[hm$Tasmanian.devil.gene.stable.ID != "",]

gt <- hm2[,2:3]
length(unique(gt$Tasmanian.devil.gene.stable.ID))
length(unique(gt$Gene.name))

```

Now run mitch for DGE2. The pathways are sourced from Reactome 7th July 2023.

```{r,mitch}

genesets <- gmt_import("../ref/ReactomePathways_2023-07-14.gmt")

m2 <- mitch_import(dge2, DEtype="deseq2", geneTable=gt)
head(m2)

res2 <- mitch_calc(m2, genesets, priority="effect",cores=16)

if ( !file.exists("mitch2.html") ) {
  mitch_report(res2, "mitch2.html")
}

top <- subset(res2$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top REACTOME Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

```

Run mitch with curated metabolic pathways.

KEGG gene sets under the "Metabolism" hierarchy were sourced from MSigDb in August 2023.

```{r, mitch2}

genesets2 <- gmt_import("../ref/gsea/gsea_KEGG.gmt")

res2 <- mitch_calc(m2, genesets2, priority="effect",cores=16)

if ( !file.exists("mitch2.html") ) {
  mitch_report(res2, "mitch2.html")
}

top <- subset(res2$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top KEGG Metabolic Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

```
Same with Reactome this time. Reactome gene sets of the metabolism hierarchy were sourced from the Reactome website in August 2023.

```{r, extract Metabolism pathways}
hierarchy <- read.table("../ref/gsea/ReactomePathwaysRelation.txt", sep ="\t", header = F, dec =".")
pathways <- read.delim("../ref/gsea/ReactomePathways.txt", header=FALSE)

hierarchy <- hierarchy[grep("HSA", hierarchy$V1), ] # keep only homo sapiens data
pathways <- pathways[grep("HSA", pathways$V1), 1:2] 

metabo_pathways <- "R-HSA-1430728" # reactome code for metabolism 

# this is very bad coding - but it works
# why 325? that is when while loop crashes because all branches of the hierarchy have been looped over
# will have to change this if Reactome data is updated
while(length(metabo_pathways) < 325){
  for(i in 1:length(metabo_pathways)){
    metabo_pathways <- unique(c(metabo_pathways,filter(hierarchy, V1 == metabo_pathways[i])$V2))
    }
  }

mpathways <- c()
for(i in 1:length(metabo_pathways)){mpathways <- c(mpathways,filter(pathways, V1 == metabo_pathways[i])$V2)}

library(magrittr)
genesets3 <- extract(genesets, mpathways)
# remove empty sets (ie hierarchy exists but no gene set is associated to it)
genesets3 <- genesets3[lapply(genesets3,length)>0]

```

```{r,mitch3}

m3 <- mitch_import(dge2, DEtype="deseq2", geneTable=gt)
res3 <- mitch_calc(m3, genesets3, priority="effect", cores=16)

if ( !file.exists("mitch_biopsy.html") ) {
  mitch_report(res3, "mitch_biopsy.html")
}

top <- subset(res3$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated metabolic pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated metabolic pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top REACTOME Metabolic Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

```