---
title: "Examining global gene expression in DFT1 and DFT2 cells"
author: "Mark Ziemann & Anne-Lise Gérard" 
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source codes: https://github.com/Anne-LiseGerard/dftd_rnaseq

```{r,pkg}

suppressPackageStartupMessages({
    library("reshape2")
    library("gplots")
    library("DESeq2")
    library("mitch")
    library("limma")
    library("kableExtra")
    library("dplyr")
    library("tidyr")
    library("ComplexHeatmap")
    library("RColorBrewer")
    library("circlize")
    library("pathview")
    library("stringr")
    library("grid")
    library("VennDiagram")
})

```

## Background

The goal of our study is to compare the expression of genes related to metabolism of DFT1 cell lines to DFT2 cell lines. 
For this, we have generated transcriptomes for 3 DFT1 and 3 DFT2 cell lines, each in triplicate.

```{r,ss1}

ss <- read.table("../ss.tsv",sep="\t",fill=TRUE,header=TRUE)
ss$DFT <- as.factor(ss$DFT)
ss$clone <- sapply(strsplit(ss$ClientID,"_"),"[[",1)

ss %>%
  kbl(caption="Sample sheet for all samples") %>%
  kable_paper("hover", full_width = F)

```

We are mainly interested in comparing all DFT1 samples against all DFT2 samples, but pairwise cell line contrasts will also be performed.
We will use Ensembl v109 for the reference transcriptome.

## Functions

```{r,func}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange,
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=1, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=1, col="red")
  mtext(SUBHEADER,cex = 0.7)
  grid()
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=1,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval")
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=1,pch=19,col="red")
}

custom_heatmap <- function(zscore, fc, aveexpr, title) {
  h1 <- Heatmap(zscore, cluster_rows = F,
              column_labels = colnames(zscore), name="Z-score",
              cluster_columns = T)
  h2 <- Heatmap(fc, row_labels = rownames(fc), row_names_gp = gpar(fontsize = 6),
              cluster_rows = F, name="logFC", top_annotation = ha, col = col_logFC,
              cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
                grid.text(round(as.numeric(fc[i, j],2)), x, y,gp=gpar(fontsize=6))
              })
  h3 <- Heatmap(aveexpr, row_labels = rownames(aveexpr), row_names_gp = gpar(fontsize = 6),
              cluster_rows = F, name = "AveExpr", col=col_AveExpr,
              cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
                grid.text(round(as.numeric(aveexpr[i, j],2)), x, y,gp=gpar(fontsize=6))},
              column_title = title)

  h <- h1 + h2 + h3
  h
}


```

## Load data

Here we load the data in from the aligner.

```{r,import}

tmp <- read.table("../fastq/3col.tsv.gz")
x <- as.data.frame(acast(tmp, V2~V1, value.var="V3",fun.aggregate=sum))
dim(x)

```

Load gene names.

```{r, genenames}

gn <- read.table("../ref/Sarcophilus_harrisii.mSarHar1.11.cdna+ncrna.genenames.tsv",fill=TRUE)

gn <- gn[order(gn$V1),]

dim(gn)

```

Load homology map

```{r,homology}

hm <- read.table("../ref/mart_export_ensembl109_2023-07-14.txt",sep="\t",header=TRUE)

```

Now need to collapse transcript data to genes.

```{r,collapse}

x$gene <- paste(gn$V2,gn$V3)

y <- aggregate(. ~ gene,x,sum)

rownames(y) <- y$gene
y$gene = NULL

dim(y)

```

## Quality control

Samples with <1M reads should be omitted.
Will also round values to integers.

```{r,qc1}

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))
barplot(cs,main="All samples",horiz=TRUE,las=1)
abline(v=1e7,col="red",lty=2)

y <- round(y)

```

## MDS

This will help us to visualise the sources of variability in the overall dataset.

Plot MDS and then remove the negative control and run Plot MDS again.

Also fix the sample names.

Plot MDS by DFT.

```{r, plotmds1}

y <- y[,colnames(y) != "DEA5-4NEG"]

ss <- ss[ss$ClientID != "DEA4_6NEG",]

colnames(y) <- sapply(strsplit(ss$ClientID,"-"),"[[",1)

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))

barplot(cs,main="All samples",horiz=TRUE,las=1)

cols <- ss$DFT
cols <- gsub("DFT1","pink",cols)
cols <- gsub("DFT2","lightblue",cols)
mymds <- plotMDS(y,plot=FALSE)

# fix the xlims
XMIN=min(mymds$x)
XMAX=max(mymds$x)
XMID=(XMAX+XMIN)/2
XMIN <- XMID + (XMIN-XMID)*1.1
XMAX <- XMID+(XMAX-XMID)*1.1

par(mar = c(5.1, 4.1, 4.1, 2.1) )
plotMDS(mymds,pch=19,cex=3,col=cols,main="MDS plot",xlim=c(XMIN,XMAX))
text(mymds,labels=colnames(y))
mtext("pink=DFT1,blue=DFT2")

y_DFT1 <- y[,colnames(y) %in% c("4906_1","4906_2","4906_3",
                                "1426_1","1426_2","1426_3",
                                "C5065_1","C5065_2","C5065_3")]
y_DFT2 <- y[,colnames(y) %in% c("RV_1","RV_2","RV_3",
                                "SN_1","SN_2","SN_3",
                                "TD549_1","TD549_2","TD549_3")]

par(mar = c(5.1, 4.1, 4.1, 2.1) )
mdsDFT1 <- plotMDS(y_DFT1,plot=FALSE)
plotMDS(mdsDFT1,pch=19,cex=3,col="pink",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT1,labels=colnames(y_DFT1))
mtext("DFT1")

par(mar = c(5.1, 4.1, 4.1, 2.1) )
mdsDFT2 <- plotMDS(y_DFT2,plot=FALSE)
plotMDS(mdsDFT2,pch=19,cex=3,col="lightblue",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT2, labels=colnames(y_DFT2))
mtext("DFT2")

```

## DESeq2

Run a simple differential analysis comparing DFT1 vs DFT2 ignoring the clone type as a source of
variation.

```{r,de1}

# split data will be necessary for the smaller comparisons (but not this one)
y1 <- y
ss1 <- ss

y1 <- y1[which(rowMeans(y1)>10),]
dim(y1)

dds <- DESeqDataSetFromMatrix(countData = y1 , colData = ss1, design = ~ DFT )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge1 <- dge
sig <- subset(dge,padj<0.05)
sig1_up <- rownames(subset(sig,log2FoldChange>0))
sig1_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig1_up)
length(sig1_dn)

```

```{r,de1viz}

maplot(dge1,"DFT1 vs DFT2")

make_volcano(dge1,"DFT1 vs DFT2")

sig[1:50,1:6] %>%
  kbl(caption="Comparison of DFT1 vs DFT2") %>%
  kable_paper("hover", full_width = F)

write.table(dge,file="dge1.tsv",sep="\t")

# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols,mar=c(5,14),
  cexRow=0.8,cexCol=0.7)

```

## Combine the three replicates

Since the previous analysis shows nearly all genes as being DE, sum replicates instead.

```{r,de2}

x4906 <- rowSums(y[,ss$clone=="4906"])
xC5065 <- rowSums(y[,ss$clone=="C5065"])
x1426 <- rowSums(y[,ss$clone=="1426"])
xRV <- rowSums(y[,ss$clone=="RV"])
xSN <- rowSums(y[,ss$clone=="SN"])
xTD549 <- rowSums(y[,ss$clone=="TD549"])

y2 <- data.frame(x4906,xC5065,x1426,xRV,xSN,xTD549)
ss2 <- as.data.frame(colnames(y2))
colnames(ss2) <- "clone"
ss2$DFT <- factor(c("DFT1","DFT1","DFT1","DFT2","DFT2","DFT2"))

y2 <- y2[which(rowMeans(y2)>10),]
dim(y2)

dds <- DESeqDataSetFromMatrix(countData = y2 , colData = ss2, design = ~ DFT )
dds2 <- dds
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge2 <- dge
sig <- subset(dge,padj<0.05)
sig2_up <- rownames(subset(sig,log2FoldChange>0))
sig2_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig2_up)
length(sig2_dn)

```

```{r,de2viz}

maplot(dge2,"DFT1 vs DFT2 aggregated")

make_volcano(dge2,"DFT1 vs DFT2 aggregated")

sig[1:50,1:6] %>%
  kbl(caption="Comparison of DFT1 vs DFT2") %>%
  kable_paper("hover", full_width = F)

write.table(dge2,file="dge2.tsv",sep="\t")

# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols[1:6],mar=c(8,14),
  cexRow=0.7,cexCol=1)

```

## Compare cell lines against one another

First 4906 vs RV, 4906 vs SN and 4906 vs TD549.
Then 1426 vs RV, 1426 vs SN and 1426 vs TD549.
Finally C5065 vs RV, C5065 vs SN and C5065 vs TD549.

```{r,de3}

yx <- vector(mode = "list", length = 9)
yx[[1]] <- select(y1, "4906_1", "4906_2", "4906_3", "RV_1", "RV_2", "RV_3")
yx[[2]] <- select(y1, "4906_1", "4906_2", "4906_3", "SN_1", "SN_2", "SN_3")
yx[[3]] <- select(y1, "4906_1", "4906_2", "4906_3", "TD549_1", "TD549_2", "TD549_3")
yx[[4]] <- select(y1, "1426_1", "1426_2", "1426_3", "RV_1", "RV_2", "RV_3")
yx[[5]] <- select(y1, "1426_1", "1426_2", "1426_3", "SN_1", "SN_2", "SN_3")
yx[[6]] <- select(y1, "1426_1", "1426_2", "1426_3", "TD549_1", "TD549_2", "TD549_3")
yx[[7]] <- select(y1, "C5065_1", "C5065_2", "C5065_3", "RV_1", "RV_2", "RV_3")
yx[[8]] <- select(y1, "C5065_1", "C5065_2", "C5065_3", "SN_1", "SN_2", "SN_3")
yx[[9]] <- select(y1, "C5065_1", "C5065_2", "C5065_3", "TD549_1", "TD549_2", "TD549_3")

ssx <- vector(mode = "list", length = 9)
dds <- vector(mode = "list", length = 9)
res <- vector(mode = "list", length = 9)
z <- vector(mode = "list", length = 9)
vsd <- vector(mode = "list", length = 9)
zz <- vector(mode = "list", length = 9)
dgex <- vector(mode = "list", length = 9)
sigx <- vector(mode = "list", length = 9)

for(i in 1:9){
  print(paste("Processing comparison: ", i, " on 9"))
  ssx[[i]] <- as.data.frame(colnames(yx[[i]]))
  colnames(ssx[[i]]) <- "clone"
  ssx[[i]]$DFT <- factor(c("DFT1","DFT1","DFT1","DFT2","DFT2","DFT2"))
  dds[[i]] <- DESeqDataSetFromMatrix(countData = yx[[i]], colData = ssx[[i]],
                                     design = ~ DFT)
  res[[i]] <- DESeq(dds[[i]], quiet=TRUE)
  z[[i]] <- results(res[[i]])
  vsd[[i]] <- vst(dds[[i]])
  zz[[i]] <- cbind(as.data.frame(z[[i]]), assay(vsd[[i]]))
  dgex[[i]] <- as.data.frame(zz[[i]][order(zz[[i]]$pvalue),])
  sigx[[i]] <- subset(dgex[[i]], padj < 0.05)
  }

plotnames <- c("4906 vs RV", "4906 vs SN", "4906 vs TD549",
               "1426 vs RV", "1426 vs SN", "1426 vs TD549",
               "C5065 vs RV", "C5065 vs SN", "C5065 vs TD549")

for(i in 1:9){
  maplot(dgex[[i]],plotnames[i])
  make_volcano(dgex[[i]],plotnames[i])
  
  sigx[[i]][1:50,1:6] %>%
  kbl(caption=plotnames[i]) %>%
  kable_paper("hover", full_width = F)
  
  write.table(dgex[[i]],file=paste(plotnames[i],".tsv"),sep="\t")
  
  mx <- sigx[[i]][,7:ncol(sigx[[i]])]
  mx <- head(mx,30)
  colfunc <- colorRampPalette(c("blue", "white", "red"))
  heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols[1:6],mar=c(8,14),
  cexRow=0.7,cexCol=1)
}

```

## Enrichment analysis

Need to get human homologs of these genes.
These were obtained from Ensembl v109 using Biomart.

```{r, hom}

rownames(dge1) <- sapply(strsplit(rownames(dge1),"\\."),"[[",1)
rownames(dge2) <- sapply(strsplit(rownames(dge2),"\\."),"[[",1)

for(i in 1:9){
  rownames(dgex[[i]]) <- sapply(strsplit(rownames(dgex[[i]]),"\\."),"[[",1)
}


hm2 <- hm[hm$Tasmanian.devil.gene.stable.ID != "",]

gt <- hm2[,2:3]
length(unique(gt$Tasmanian.devil.gene.stable.ID))
length(unique(gt$Gene.name))

for(i in 1:length(gt$Gene.name)){
  if(gt$Gene.name[i]==""){gt$Gene.name[i] <- gt$Tasmanian.devil.gene.stable.ID[i]}
}

```

Now run mitch for DGE2. The pathways were sourced from Reactome on 7th July 2023.

```{r,mitch}

genesets <- gmt_import("../ref/ReactomePathways_2023-07-14.gmt")

m2 <- mitch_import(dge2, DEtype="deseq2", geneTable=gt)
head(m2)

res2 <- mitch_calc(m2, genesets, priority="effect",cores=16)

if ( !file.exists("mitch2.html") ) {
  mitch_report(res2, "mitch2.html")
}

top <- subset(res2$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top REACTOME Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

react_metabosets <- readRDS("../ref/reactome_metabo.rds")

top_metabo <- filter(top, set %in% names(react_metabosets))

barplot(rev(abs(top_metabo$s.dist)),horiz=TRUE,las=1,names.arg=rev(top_metabo$set),
  main="Enriched Metabolic Pathways - REACTOME",cex.names=0.6,xlab="Enrichment Score",
  col=c("red","blue","red","blue"),
  xlim=c(0,0.8))

```
KEGG gene sets were sourced from MSigDb on 14 September 2023.

```{r,mitch2}

genesets2 <- gmt_import("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt.txt")

res2 <- mitch_calc(m2, genesets2, priority="effect",cores=16)

if ( !file.exists("mitch2.html") ) {
  mitch_report(res2, "mitch2.html")
}

top <- subset(res2$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top KEGG Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

kegg_metabosets <- gmt_import("../ref/gsea/gsea_KEGG.gmt")

top_metabo <- filter(top, set %in% names(kegg_metabosets))

barplot(rev(abs(top_metabo$s.dist)),horiz=TRUE,las=1,names.arg=rev(top_metabo$set),
  main="Enriched Metabolic Pathways - KEGG",cex.names=0.6,xlab="Enrichment Score",
  col=c("red"),
  xlim=c(0,0.8))

```
## Visualisation

Heatmaps.
```{r, complexheatmap1}

# see https://www.youtube.com/watch?v=ht1r34-ifVI for reference

# find corresponding human genes to devil ENS IDs
dge3 <- dge2 
dge3$ENSSHAG <- rownames(dge3)
hm3 <- data.frame(ENSSHAG=hm2[,2],geneID=hm2[,3])
dge3 <- left_join(dge3, hm3, by="ENSSHAG")
dge3 <- unite(dge3,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ", remove=F)
dge3 <- unique(dge3)
rownames(dge3) <- dge3$ENSSHAG_geneID

# only plot significant DEGs
dge3sig <- filter(dge3, padj < 0.05)
dge3sig <- dge3sig[order(dge3sig$log2FoldChange, decreasing=TRUE),]

mx <- dge3sig[,7:ncol(dge3sig)] # get normalised counts
 
mx.scaled <- t(apply(mx[,1:6], 1, scale)) # center and scale each column (Z-score)
colnames(mx.scaled) <- colnames(mx[,1:6])
log2FC <- as.matrix(dge3sig$log2FoldChange)
rownames(log2FC) <- rownames(dge3sig)
colnames(log2FC) <- "logFC"
mean <- as.matrix(dge3sig$baseMean)
rownames(mean) <- rownames(dge3sig)
colnames(mean) <- "AveExpr"

col_logFC <- colorRamp2(c(min(log2FC),0,max(log2FC)), c("blue","white","red"))
col_AveExpr <- colorRamp2(c(quantile(mean)[1], quantile(mean)[4]), c("white","red"))

ha <- HeatmapAnnotation(summary=anno_summary(gp=gpar(fill=2),height=unit(2, "cm")))

# ----- Top DEGs in REACTOME metabolic pathways ------

metabo <- c() # get all names of genes involved in metabolism
for(i in 1:length(genesets)){metabo <- c(metabo,genesets[[i]])}
metabo <- unique(metabo) # get rid of duplicates

temp <- c() # get ENSSHAG and gene ID
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

# top 20 up & down
met40 <- rbind(head(met,20),tail(met,20))
met40FC <- as.matrix(c(head(metFC,20),tail(metFC,20)))
rownames(met40FC) <- rownames(met40)
colnames(met40FC) <- "logFC"
met40AE <- as.matrix(c(head(metAE,20),tail(metAE,20)))
rownames(met40AE) <- rownames(met40)
colnames(met40AE) <- "AveExpr"

custom_heatmap(met40, met40FC, met40AE, title = "TOP 40 - REACTOME metabolism")

# ----- Degradation of cysteine and homocysteine -----

metabo <- c() # get all names of genes involved in metabolism
for(i in 1:length(genesets)){metabo <- c(metabo,genesets[["Degradation of cysteine and homocysteine"]])}
metabo <- unique(metabo) # get rid of duplicates

temp <- c() # get ENSSHAG and gene ID
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

custom_heatmap(met, metFC, metAE, title = "Degradation of cysteine and homocysteine - REACTOME")

# ----- Cholesterol biosynthesis -----

metabo <- c() # get all names of genes involved in metabolism
for(i in 1:length(genesets)){metabo <- c(metabo,genesets[["Cholesterol biosynthesis"]])}
metabo <- unique(metabo) # get rid of duplicates

temp <- c() # get ENSSHAG and gene ID
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

custom_heatmap(met, metFC, metAE, title = "Cholesterol biosynthesis - REACTOME")

# ----- Sulfur amino acid metabolism -----

metabo <- c() # get all names of genes involved in metabolism
for(i in 1:length(genesets)){metabo <- c(metabo,genesets[["Sulfur amino acid metabolism"]])}
metabo <- unique(metabo) # get rid of duplicates

temp <- c() # get ENSSHAG and gene ID
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

custom_heatmap(met, metFC, metAE, title = "Sulfur amino acid metabolism - REACTOME")

# ----- Selenoamino acid metabolism -----

metabo <- c() # get all names of genes involved in metabolism
for(i in 1:length(genesets)){metabo <- c(metabo,genesets[["Selenoamino acid metabolism"]])}
metabo <- unique(metabo) # get rid of duplicates

temp <- c() # get ENSSHAG and gene ID
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

custom_heatmap(met, metFC, metAE, title = "Selenoamino acid metabolism - REACTOME")

# ----- Top DEGs in KEGG metabolic pathways ------

metabo <- c()
for(i in 1:length(genesets2)){metabo <- c(metabo,genesets2[[i]])}
metabo <- unique(metabo)

temp <- c()
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

# top 20 up & down
met40 <- rbind(head(met,20),tail(met,20))
met40FC <- as.matrix(c(head(metFC,20),tail(metFC,20)))
rownames(met40FC) <- rownames(met40)
colnames(met40FC) <- "logFC"
met40AE <- as.matrix(c(head(metAE,20),tail(metAE,20)))
rownames(met40AE) <- rownames(met40)
colnames(met40AE) <- "AveExpr"

custom_heatmap(met40, met40FC, met40AE, title = "TOP 40 - KEGG metabolism")

# ----- Steroid biosynthesis -----

metabo <- c() # get all names of genes involved in metabolism
for(i in 1:length(genesets2)){metabo <- c(metabo,genesets2[["KEGG_STEROID_BIOSYNTHESIS"]])}
metabo <- unique(metabo) # get rid of duplicates

temp <- c() # get ENSSHAG and gene ID
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

custom_heatmap(met, metFC, metAE, title = "Steroid biosynthesis - KEGG")
```

Venn diagrams data of all DEGs, and DEGs linked to metabolism.
```{r, Venn data}

alldegs <- rownames(dge3sig)
updegs <- rownames(filter(dge3sig, log2FoldChange > 0))
downdegs <- rownames(filter(dge3sig, log2FoldChange < 0))
metdegs <- rownames(met)

venn_clines <- list(alldegs, updegs, downdegs, metdegs)
saveRDS(venn_clines, file = "venn_clines.rds")

# venn.diagram(
#   x = list(alldegs, updegs, downdegs),
#   category.names = c("All DEGs" , "Up", "Down"),
#   filename = 'updown_venn_diagramm.png',
#   output=TRUE
# )

```

Pathview to visualise metabolic pathways. 

```{r, pathview1}

pathFC <- log2FC
rownames(pathFC) <- str_split(rownames(pathFC), " ", simplify=T)[,2]

# cholesterol
pathview(gene.data=pathFC, pathway.id="hsa04979", species="hsa",gene.idtype="SYMBOL", node.sum="sum",
         limit=list(gene=c(-10,10), cpd=1), bins=list(gene=20, cpd=1),
         low="blue",mid="white",high="red",
         same.layer=FALSE)
# cysteine and methionine
pathview(gene.data=pathFC, pathway.id="hsa00270", species="hsa",gene.idtype="SYMBOL", node.sum="sum",
         limit=list(gene=c(-10,10), cpd=1), bins=list(gene=20, cpd=1),
         low="blue",mid="white",high="red",
         same.layer=FALSE)
# steroid biosynthesis
pathview(gene.data=pathFC, pathway.id="hsa00100", species="hsa",gene.idtype="SYMBOL", node.sum="sum",
         limit=list(gene=c(-10,10), cpd=1), bins=list(gene=20, cpd=1),
         low="blue",mid="white",high="red",
         same.layer=FALSE)
```
```{r SBGNview}
# library(SBGNview)
# input.pathways <- findPathways("Cholesterol biosynthesis")
# SBGNview.obj <- SBGNview(
#           gene.data = pathFC, 
#           gene.id.type = "SYMBOL",
#           input.sbgn = input.pathways[7,]$pathway.id,
#           output.file = "quick.start", 
#           output.formats =  c("png")
#           ) 
# print(SBGNview.obj)
```

Run mitch on cell line pairwise comparisons.

```{r, mitchx}

mx <- vector(mode = "list", length = 9)
resx <- vector(mode = "list", length = 9)

for(i in 1:9){
  
  mx[[i]] <- mitch_import(dgex[[i]], DEtype="deseq2",geneTable=gt)
  resx[[i]] <- mitch_calc(mx[[i]], genesets3, priority="effect",cores=16)

  if ( !file.exists(paste(plotnames[i],"_mitch.html")) ) {
    mitch_report(resx[[i]], paste(plotnames[i],"_mitch.html"))
  }
  
  top <- subset(resx[[i]]$enrichment_result, p.adjustANOVA<0.05)
  
  topup <- head(subset(top,s.dist>0),20)
  
  topup %>%
  kbl(caption=paste("Top 20 upregulated metabolic pathways: ", plotnames[i])) %>%
  kable_paper("hover", full_width = F)
  
  topdn <- head(subset(top,s.dist<0),20)
  
  topdn %>%
    kbl(caption=paste("Top 20 downregulated metabolic pathways: ", plotnames[i])) %>%
    kable_paper("hover", full_width = F)
  
  top2 <- rbind(head(topup,10),head(topdn,10))
  top2 <- top2[order(-top2$s.dist),]
  par(mar=c(5,15,5,2))
  
  barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
          main=paste(plotnames[i]),
          cex.names=0.6,xlab="Enrichment Score",
          col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
          xlim=c(0,0.8))
  
}

```

Make heatmaps for pairwise cell line comparisons.

```{r, complexheatmap2}

# ----- Fatty acid metabolism heatmaps -----
# fatty1 <- genesets2[["REACTOME_FATTY_ACID_METABOLISM"]]
# fatty2 <- genesets2[["HALLMARK_FATTY_ACID_METABOLISM"]]
# fatty3 <- genesets2[["GOBP_FATTY_ACID_BIOSYNTHETIC_PROCESS"]]
# fatty4 <-genesets2[["KEGG_FATTY_ACID_METABOLISM"]]
# fatty <- list(fatty1=fatty1,fatty2=fatty2,fatty3=fatty3,fatty4=fatty4)
# title <- c("REACTOME_FATTY_ACID_METABOLISM","HALLMARK_FATTY_ACID_METABOLISM","GOBP_FATTY_ACID_BIOSYNTHETIC_PROCESS","KEGG_FATTY_ACID_METABOLISM")
# 
# for(i in 1:length(fatty)){
#   temp <- c()
#   for(j in 1:length(fatty[[i]])){temp <- rbind(temp,filter(hm3, geneID==fatty[[i]][j]))}
#   fattyx <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))
#   fat <- subset(mx.scaled, rownames(mx.scaled) %in% fattyx)
#   fatFC <- subset(log2FC, rownames(log2FC) %in% fattyx)
#   fatAE <- subset(mean, rownames(mean) %in% fattyx)
#   
#   h1 <- Heatmap(fat, cluster_rows = F,
#                 column_labels = colnames(fat), name="Z-score",
#                 cluster_columns = T)
#   h2 <- Heatmap(fatFC, row_labels = rownames(fatFC), row_names_gp = gpar(fontsize = 6),
#               cluster_rows = F, name="logFC", top_annotation = ha, col = col_logFC,
#               cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
#                 grid.text(round(as.numeric(fatFC[i, j],2)), x, y, gp=gpar(fontsize=6))
#               })
#   h3 <- Heatmap(fatAE, row_labels = rownames(fatAE), row_names_gp = gpar(fontsize = 6),
#               cluster_rows = F, name = "AveExpr", col=col_AveExpr,
#               cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
#                 grid.text(round(as.numeric(fatAE[i, j],2)), x, y, gp=gpar(fontsize=6))},
#               column_title=title[i])
#   
#   name <- paste("q",i,sep="")
#   assign(name, h1 + h2 + h3)
# }
# 
# q1
# q2
# q3
# q4

```

Pathview to visualise metabolic pathways. 

```{r, pathview2}


```


## Session information

For reproducibility.

```{r,session}

sessionInfo()

```
