---
title: "Examining global gene expression in DFT1 and DFT2 cells"
author: "Mark Ziemann & Anne-Lise Gérard" 
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source codes: https://github.com/Anne-LiseGerard/dftd_rnaseq

```{r,pkg}

suppressPackageStartupMessages({
    library("reshape2")
    library("gplots")
    library("DESeq2")
    library("mitch")
    library("limma")
    library("kableExtra")
    library("dplyr")
})

```

## Background

The goal of our study is to compare the expression of genes related to metabolism of DFT1 cell lines to
DFT2 cell lines.
Our main goal is comparing genes expressed in the following pathways: Glycolysis, Pentose Phosphate
Pathway, Fatty Acid Metabolism, Glutaminolysis and Oxidative Phosphorylation.
A secondary goal would be to compare the expression of genes related to the production of Reactive
Oxygen Species (ROS) and protection against ROS (genes involved in DNA repair, production of
antioxydants etc).
Finally, we might also be interested in looking into genes linked to cholesterol metabolism as this has
been shown to be important for DFT1 but I need to do some additional reading to know what exactly we're
after.
See sample sheet below (edited from Yin Peng), there are 3 DFT1 and 3 DFT2 cell lines, and 3 replicates
of each cell line:

```{r,ss1}

ss <- read.table("../ss.tsv",sep="\t",fill=TRUE,header=TRUE)
ss$DFT <- as.factor(ss$DFT)
ss$clone <- sapply(strsplit(ss$ClientID,"_"),"[[",1)

ss %>%
  kbl(caption="Sample sheet for all samples") %>%
  kable_paper("hover", full_width = F)

```

We are mainly interested in comparing all DFT1 samples against all DFT2 samples.
However, we did notice that some cell lines within a same DFT produced different results in the
metabolism experiments so we might need to compare all 6 cell lines against one another and not be able to simply do a DFT1 vs DFT2 contrast, if that makes sense.

Regarding the reference transcriptome, we will use Ensembl v109.

## Functions

```{r,func}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange,
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=1, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=1, col="red")
  mtext(SUBHEADER,cex = 0.7)
  grid()
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=1,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval")
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=1,pch=19,col="red")
}

```

## Load data

Here we load the data in from the aligner.

```{r,import}

tmp <- read.table("../fastq/3col.tsv.gz")
x <- as.data.frame(acast(tmp, V2~V1, value.var="V3",fun.aggregate=sum))
dim(x)

```

Load gene names.

```{r,genenames}

gn <- read.table("../ref/Sarcophilus_harrisii.mSarHar1.11.cdna+ncrna.genenames.tsv",fill=TRUE)

gn <- gn[order(gn$V1),]

dim(gn)

```

Load homology map

```{r,homology}

hm <- read.table("../ref/mart_export_ensembl109_2023-07-14.txt",sep="\t",header=TRUE)

```

Now need to collapse transcript data to genes.

```{r,collapse}

x$gene <- paste(gn$V2,gn$V3)

y <- aggregate(. ~ gene,x,sum)

rownames(y) <- y$gene
y$gene = NULL

dim(y)

```

## Quality control

Samples with <1M reads should be omitted.
Will also round values to integers.

```{r,qc1}

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))
barplot(cs,main="All samples",horiz=TRUE,las=1)
abline(v=1e7,col="red",lty=2)

y <- round(y)

```

## MDS

This will help us to visualise the sources of variability in the overall dataset.

Plot MDS and then remove the negative control and run Plot MDS again.

Also fix the sample names.

Plot MDS by DFT.

```{r,plotmds1}

par(mar = c(5.1, 4.1, 4.1, 2.1) )

plotMDS(y)

y <- y[,colnames(y) != "DEA5-4NEG"]

plotMDS(y)

ss <- ss[ss$ClientID != "DEA4_6NEG",]

colnames(y) <- sapply(strsplit(ss$ClientID,"-"),"[[",1)

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))
barplot(cs,main="All samples",horiz=TRUE,las=1)

par(mar = c(5.1, 4.1, 4.1, 2.1) )

plotMDS(y)

cols <- ss$DFT
cols <- gsub("DFT1","pink",cols)
cols <- gsub("DFT2","lightblue",cols)
mymds <- plotMDS(y,plot=FALSE)

# fix the xlims
XMIN=min(mymds$x)
XMAX=max(mymds$x)
XMID=(XMAX+XMIN)/2
XMIN <- XMID + (XMIN-XMID)*1.1
XMAX <- XMID+(XMAX-XMID)*1.1
plotMDS(mymds,pch=19,cex=3,col=cols,main="MDS plot",xlim=c(XMIN,XMAX))
text(mymds,labels=colnames(y))
mtext("pink=DFT1,blue=DFT2")


y_DFT1 <- y[,colnames(y) %in% c("4906_1","4906_2","4906_3",
                                "1426_1","1426_2","1426_3",
                                "C5065_1","C5065_2","C5065_3")]
y_DFT2 <- y[,colnames(y) %in% c("RV_1","RV_2","RV_3",
                                "SN_1","SN_2","SN_3",
                                "TD549_1","TD549_2","TD549_3")]

mdsDFT1 <- plotMDS(y_DFT1,plot=FALSE)
plotMDS(mdsDFT1,pch=19,cex=3,col="pink",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT1,labels=colnames(y_DFT1))
mtext("DFT1")

mdsDFT2 <- plotMDS(y_DFT2,plot=FALSE)
plotMDS(mdsDFT2,pch=19,cex=3,col="lightblue",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT2, labels=colnames(y_DFT2))
mtext("DFT2")

```

## DESeq2

Run a simple differential analysis comparing DFT1 vs DFT2 ignoring the clone type as a source of
variation.

```{r,de1}

# split data will be necessary for the smaller comparisons (but not this one)
y1 <- y
ss1 <- ss

y1 <- y1[which(rowMeans(y1)>10),]
dim(y1)

dds <- DESeqDataSetFromMatrix(countData = y1 , colData = ss1, design = ~ DFT )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge1 <- dge
sig <- subset(dge,padj<0.05)
sig1_up <- rownames(subset(sig,log2FoldChange>0))
sig1_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig1_up)
length(sig1_dn)

```

```{r,de1viz}

maplot(dge1,"DFT1 vs DFT2")

make_volcano(dge1,"DFT1 vs DFT2")

sig[1:50,1:6] %>%
  kbl(caption="Comparison of DFT1 vs DFT2") %>%
  kable_paper("hover", full_width = F)

write.table(dge,file="dge1.tsv",sep="\t")

# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols,mar=c(5,14),
  cexRow=0.8,cexCol=0.7)

```

## Combine the three replicates

Sum replicates.

```{r,de2}

x4906 <- rowSums(y[,ss$clone=="4906"])
xC5065 <- rowSums(y[,ss$clone=="C5065"])
x1426 <- rowSums(y[,ss$clone=="1426"])
xRV <- rowSums(y[,ss$clone=="RV"])
xSN <- rowSums(y[,ss$clone=="SN"])
xTD549 <- rowSums(y[,ss$clone=="TD549"])

y2 <- data.frame(x4906,xC5065,x1426,xRV,xSN,xTD549)
ss2 <- as.data.frame(colnames(y2))
colnames(ss2) <- "clone"
ss2$DFT <- factor(c("DFT1","DFT1","DFT1","DFT2","DFT2","DFT2"))

y2 <- y2[which(rowMeans(y2)>10),]
dim(y2)

dds <- DESeqDataSetFromMatrix(countData = y2 , colData = ss2, design = ~ DFT )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge2 <- dge
sig <- subset(dge,padj<0.05)
sig2_up <- rownames(subset(sig,log2FoldChange>0))
sig2_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig2_up)
length(sig2_dn)

```

```{r,de2viz}

maplot(dge2,"DFT1 vs DFT2 aggregated")

make_volcano(dge2,"DFT1 vs DFT2 aggregated")

sig[1:50,1:6] %>%
  kbl(caption="Comparison of DFT1 vs DFT2") %>%
  kable_paper("hover", full_width = F)

write.table(dge,file="dge2.tsv",sep="\t")

# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols[1:6],mar=c(8,14),
  cexRow=0.7,cexCol=1)

```

## Compare cell lines against one another

First 4906 vs RV, 4906 vs SN and 4906 vs TD549.
Then 1426 vs RV, 1426 vs SN and 1426 vs TD549.
Finally C5065 vs RV, C5065 vs SN and C5065 vs TD549.

```{r,de3}

#TODO: make loops to perform all comparisons
y3 <- select(y1, "4906_1", "4906_2", "4906_3", "RV_1", "RV_2", "RV_3")
y4 <- select(y1, "4906_1", "4906_2", "4906_3", "SN_1", "SN_2", "SN_3")
y5 <- select(y1, "4906_1", "4906_2", "4906_3", "TD549_1", "TD549_2", "TD549_3")

ss3 <- as.data.frame(colnames(y3))
colnames(ss3) <- "clone"
ss3$DFT <- factor(c("DFT1","DFT1","DFT1","DFT2","DFT2","DFT2"))

dds <- DESeqDataSetFromMatrix(countData = y3, colData = ss3, design = ~ DFT )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge3<-as.data.frame(zz[order(zz$pvalue),])
sig <- subset(dge3,padj<0.05)
sig2_up <- rownames(subset(sig,log2FoldChange>0))
sig2_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig2_up)
length(sig2_dn)

maplot(dge3,"4906 vs RV")

make_volcano(dge3,"4906 vs RV")

sig[1:50,1:6] %>%
  kbl(caption="Comparison of 4906 vs RV") %>%
  kable_paper("hover", full_width = F)

write.table(dge,file="dge3.tsv",sep="\t")

# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols[1:6],mar=c(8,14),
  cexRow=0.7,cexCol=1)

```

## Enrichment analysis

Need to get human homologs of these genes.
These were obtained from Ensembl v109 biomart (website).

```{r}

rownames(dge2) <- sapply(strsplit(rownames(dge2),"\\."),"[[",1)
rownames(dge3) <- sapply(strsplit(rownames(dge3),"\\."),"[[",1)

hm2 <- hm[hm$Tasmanian.devil.gene.stable.ID != "",]

gt <- hm2[,2:3]
length(unique(gt$Tasmanian.devil.gene.stable.ID))
length(unique(gt$Gene.name))

```

Now run mitch for DGE2.
There will be a report generated that has more details on the enrichment analysis.
The pathways are sourced from Reactome 7th July 2023.

```{r,mitch2}

genesets <- gmt_import("../ref/ReactomePathways_2023-07-14.gmt")

m2 <- mitch_import(dge2, DEtype="deseq2",geneTable=gt)
head(m2)

res2 <- mitch_calc(m2, genesets, priority="effect",cores=16)

if ( !file.exists("mitch2.html") ) {
  mitch_report(res2, "mitch2.html")
}

top <- subset(res2$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

```

Run mitch with curated metabolic pathways.
These pathways were sourced from MSigDb on 31/07/2023.

```{r,mitch3}

genesets2 <- gmt_import("../ref/gsea/gsea.gmt")

m3 <- mitch_import(dge2, DEtype="deseq2",geneTable=gt)

res3 <- mitch_calc(m3, genesets2, priority="effect",cores=16)

if ( !file.exists("mitch3.html") ) {
  mitch_report(res4, "mitch3.html")
}

top <- subset(res3$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated metabolic pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated metabolic pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top Metabolic Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

```

Run mitch on cell line pairwise comparisons

```{r,mitch4}

m4 <- mitch_import(dge3, DEtype="deseq2",geneTable=gt)

res4 <- mitch_calc(m4, genesets2, priority="effect",cores=16)

if ( !file.exists("mitch4.html") ) {
  mitch_report(res4, "mitch4.html")
}

top <- subset(res4$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated metabolic pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated metabolic pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top Metabolic Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

```


## Session information

For reproducibility.

```{r,session}

sessionInfo()

```
