---
title: "Examining global gene expression in DFT1 and DFT2 cells"
author: "Mark Ziemann & Anne-Lise Gérard"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source codes: https://github.com/Anne-LiseGerard/dftd_rnaseq

```{r, pkg}

suppressPackageStartupMessages({
    library("reshape2")
    library("gplots")
    library("DESeq2")
    library("mitch")
    library("limma")
    library("kableExtra")
    library("dplyr")
    library("tidyr")
    library("ComplexHeatmap")
    library("RColorBrewer")
    library("circlize")
    library("pathview")
    library("stringr")
    library("grid")
    library("VennDiagram")
    library("enrichplot")
    library("ggplot2")
    library("viridis")
    library("SBGNview")
    library("readxl")
    library("cowplot")
})

knitr::opts_chunk$set(dev = 'svg')

```

## Background

The goal of our study is to compare the expression of genes related to metabolism of DFT1 cell lines to DFT2 cell lines. 
For this, we have generated transcriptomes for 3 DFT1 and 3 DFT2 cell lines, each in triplicate.

```{r,ss1}

ss <- read.table("../ss.tsv",sep="\t",fill=TRUE,header=TRUE)
ss$DFT <- as.factor(ss$DFT)
ss$clone <- sapply(strsplit(ss$ClientID,"_"),"[[",1)

ss %>%
  kbl(caption="Sample sheet for all samples") %>%
  kable_paper("hover", full_width = F)

```

We are mainly interested in comparing all DFT1 samples against all DFT2 samples.
We will use mSarHar 1.11 from Ensembl v109 for the reference transcriptome.

## Functions

```{r,func}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange,
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=1, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=1, col="#5297A7")
  mtext(SUBHEADER,cex = 0.7)
  grid()
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=1,pch=19,col="#D5D7E2",
        main=name, xlab="log2 FC", ylab="-log10 pval")
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=1,pch=19,col="#5297A7")
}

custom_heatmap <- function(zscore, fc, aveexpr, title) {
  h1 <- Heatmap(zscore, cluster_rows = F,
              column_labels = colnames(zscore), name="Z-score",
              cluster_columns = T, 
              col= colorRamp2(c(min(zscore),0,max(zscore)), hcl_palette = "Blue-Red 2"))
  h2 <- Heatmap(fc, row_labels = rownames(fc), row_names_gp = gpar(fontsize = 6),
              cluster_rows = F, name="logFC", top_annotation = ha, col = col_logFC,
              cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
                grid.text(round(as.numeric(fc[i, j],2)), x, y,gp=gpar(fontsize=6))
              }, column_title = title)
  
  # h3 <- Heatmap(aveexpr, row_labels = rownames(aveexpr), row_names_gp = gpar(fontsize = 6),
  #             cluster_rows = F, name = "AveExpr", col=col_AveExpr,
  #             cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
  #               grid.text(round(as.numeric(aveexpr[i, j],2)), x, y,gp=gpar(fontsize=6))},
  #             column_title = title)

  h <- h1 + h2 #+ h3
  h
}


```

## Load data

Here we load the data in from the aligner.

```{r,import}

tmp <- read.table("../fastq/3col.tsv.gz")
x <- as.data.frame(acast(tmp, V2~V1, value.var="V3",fun.aggregate=sum))
dim(x)

```

Load gene names.

```{r, genenames}

gn <- read.table("../ref/Sarcophilus_harrisii.mSarHar1.11.cdna+ncrna.genenames.tsv",fill=TRUE)

gn <- gn[order(gn$V1),]

dim(gn)

```

Load homology map

```{r,homology}

hm <- read.table("../ref/mart_export_ensembl109_2023-07-14.txt",sep="\t",header=TRUE)

```

Now need to collapse transcript data to genes.

```{r,collapse}

x$gene <- paste(gn$V2,gn$V3)

y <- aggregate(. ~ gene,x,sum)

rownames(y) <- y$gene
y$gene = NULL

dim(y)

```

## Quality control

Samples with <1M reads should be omitted.
Will also round values to integers.

```{r,qc1}

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))
barplot(cs,main="All samples",horiz=TRUE,las=1)
abline(v=1e7,col="red",lty=2)

y <- round(y)

```

## MDS

This will help us to visualise the sources of variability in the overall dataset.

Plot MDS and then remove the negative control and run Plot MDS again.

Also fix the sample names.

Plot MDS by DFT.

```{r, plotmds1}

y <- y[,colnames(y) != "DEA5-4NEG"]

ss <- ss[ss$ClientID != "DEA4_6NEG",]

colnames(y) <- sapply(strsplit(ss$ClientID,"-"),"[[",1)

saveRDS(y, file = "y_clines.rds")

cs <- colSums(y)
cs <- cs[order(cs)]

par(mar=c(5,10,5,2))

barplot(cs,main="All samples",horiz=TRUE,las=1, xlab="reads")

cols <- ss$DFT
cols <- gsub("DFT1","#B3B9DF",cols)
cols <- gsub("DFT2","#DDAFB7",cols)
mymds <- plotMDS(y,plot=FALSE)

# fix the xlims
XMIN=min(mymds$x)
XMAX=max(mymds$x)
XMID=(XMAX+XMIN)/2
XMIN <- XMID + (XMIN-XMID)*1.1
XMAX <- XMID+(XMAX-XMID)*1.1
par(mar = c(5.1, 4.1, 4.1, 2.1) )
plotMDS(mymds,pch=19,cex=3,col=cols,main="Cell lines",xlim=c(XMIN,XMAX))
text(mymds,labels=colnames(y))
mtext("blue=DFT1, pink=DFT2")

y_DFT1 <- y[,colnames(y) %in% c("4906_1","4906_2","4906_3",
                                "1426_1","1426_2","1426_3",
                                "C5065_1","C5065_2","C5065_3")]
y_DFT2 <- y[,colnames(y) %in% c("RV_1","RV_2","RV_3",
                                "SN_1","SN_2","SN_3",
                                "TD549_1","TD549_2","TD549_3")]

par(mar = c(5.1, 4.1, 4.1, 2.1) )
mdsDFT1 <- plotMDS(y_DFT1,plot=FALSE)
plotMDS(mdsDFT1,pch=19,cex=3,col="#B3B9DF",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT1,labels=colnames(y_DFT1))
mtext("DFT1")

par(mar = c(5.1, 4.1, 4.1, 2.1) )
mdsDFT2 <- plotMDS(y_DFT2,plot=FALSE)
plotMDS(mdsDFT2,pch=19,cex=3,col="#DDAFB7",main="MDS plot",xlim=c(XMIN,XMAX))
text(mdsDFT2, labels=colnames(y_DFT2))
mtext("DFT2")

```

## DESeq2

Run a simple differential analysis comparing DFT1 vs DFT2 ignoring the clone type as a source of
variation.

```{r,de1}

# split data will be necessary for the smaller comparisons (but not this one)
y1 <- y
ss1 <- ss

y1 <- y1[which(rowMeans(y1)>10),] # remove extremely low expression genes
dim(y1)

dds <- DESeqDataSetFromMatrix(countData = y1 , colData = ss1, design = ~ DFT )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge1 <- dge
sig <- subset(dge,padj<0.05)
sig1_up <- rownames(subset(sig,log2FoldChange>0))
sig1_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig1_up)
length(sig1_dn)

```

```{r,de1viz}

maplot(dge1,"DFT1 vs DFT2")

make_volcano(dge1,"DFT1 vs DFT2")

sig[1:50,1:6] %>%
  kbl(caption="Comparison of DFT1 vs DFT2") %>%
  kable_paper("hover", full_width = F)

write.table(dge,file="dge1.tsv",sep="\t")

# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols,mar=c(5,14),
  cexRow=0.8,cexCol=0.7)

```

## Combine the three replicates

Since the previous analysis shows nearly all genes as being DE, sum replicates instead.

```{r,de2}

x4906 <- rowSums(y[,ss$clone=="4906"])
xC5065 <- rowSums(y[,ss$clone=="C5065"])
x1426 <- rowSums(y[,ss$clone=="1426"])
xRV <- rowSums(y[,ss$clone=="RV"])
xSN <- rowSums(y[,ss$clone=="SN"])
xTD549 <- rowSums(y[,ss$clone=="TD549"])

y2 <- data.frame(x4906,xC5065,x1426,xRV,xSN,xTD549)
ss2 <- as.data.frame(colnames(y2))
colnames(ss2) <- "clone"
ss2$DFT <- factor(c("DFT1","DFT1","DFT1","DFT2","DFT2","DFT2"))

y2 <- y2[which(rowMeans(y2)>10),]
dim(y2)

dds <- DESeqDataSetFromMatrix(countData = y2 , colData = ss2, design = ~ DFT )
dds2 <- dds
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
dge2 <- dge
sig <- subset(dge,padj<0.05)
sig2 <- sig
sig2_up <- rownames(subset(sig,log2FoldChange>0))
sig2_dn <- rownames(subset(sig,log2FoldChange<0))
length(sig2_up)
length(sig2_dn)

```

```{r,de2viz}

maplot(dge2,"Cell lines")

make_volcano(dge2,"Cell lines")

sig[1:50,1:6] %>%
  kbl(caption="Comparison of DFT1 vs DFT2") %>%
  kable_paper("hover", full_width = F)

write.table(dge2,file="dge2.tsv",sep="\t")

```

```{r,de2vizheatmap,fig.dim = c(4, 6)}
# heatmap
mx <- sig[,7:ncol(sig)]
mx <- head(mx,30)
colfunc <- colorRampPalette(viridis(256))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),ColSideColors=cols[1:6],mar=c(8,14),
  cexRow=0.7,cexCol=1)

# top 20 up & top 20 down
sig <- sig[order(sig$log2FoldChange, decreasing=TRUE),]
sig_noNA <- sig
sig_noNA$genes <- rownames(sig_noNA)
sig_noNA <- filter(sig_noNA, !grepl(' NA', genes))

top <- rbind(head(sig_noNA,20),tail(sig_noNA,20))

mx <- top[,7:ncol(top)] # get normalised counts

mx.scaled <- t(apply(mx[,1:6], 1, scale)) # center and scale each column (Z-score)
colnames(mx.scaled) <- colnames(mx[,1:6])
colnames(mx.scaled) <- c("DFT1 4906", "DFT1 C5065", "DFT1 1426", "DFT2 RV", "DFT2 SN", "DFT2 TD549")
log2FC <- as.matrix(top$log2FoldChange)
rownames(log2FC) <- rownames(top)
colnames(log2FC) <- "logFC"
mean <- as.matrix(top$baseMean)
rownames(mean) <- rownames(top)
colnames(mean) <- "AveExpr"

col_logFC <- colorRamp2(c(min(log2FC),0,max(log2FC)), hcl_palette = "Blue-Red 2")
#col_AveExpr <- colorRamp2(c(quantile(mean)[1], quantile(mean)[4]), c("white","red"))

ha <- HeatmapAnnotation(summary=anno_summary(gp=gpar(fill="#D5D7E2"), height=unit(2, "cm")))

custom_heatmap(mx.scaled, log2FC, mean, title = "Top 40 DEGs - Cell lines")

```

## Compare cell lines against one another

First 4906 vs RV, 4906 vs SN and 4906 vs TD549.
Then 1426 vs RV, 1426 vs SN and 1426 vs TD549.
Finally C5065 vs RV, C5065 vs SN and C5065 vs TD549.

```{r,de3}
# 
# yx <- vector(mode = "list", length = 9)
# yx[[1]] <- select(y1, "4906_1", "4906_2", "4906_3", "RV_1", "RV_2", "RV_3")
# yx[[2]] <- select(y1, "4906_1", "4906_2", "4906_3", "SN_1", "SN_2", "SN_3")
# yx[[3]] <- select(y1, "4906_1", "4906_2", "4906_3", "TD549_1", "TD549_2", "TD549_3")
# yx[[4]] <- select(y1, "1426_1", "1426_2", "1426_3", "RV_1", "RV_2", "RV_3")
# yx[[5]] <- select(y1, "1426_1", "1426_2", "1426_3", "SN_1", "SN_2", "SN_3")
# yx[[6]] <- select(y1, "1426_1", "1426_2", "1426_3", "TD549_1", "TD549_2", "TD549_3")
# yx[[7]] <- select(y1, "C5065_1", "C5065_2", "C5065_3", "RV_1", "RV_2", "RV_3")
# yx[[8]] <- select(y1, "C5065_1", "C5065_2", "C5065_3", "SN_1", "SN_2", "SN_3")
# yx[[9]] <- select(y1, "C5065_1", "C5065_2", "C5065_3", "TD549_1", "TD549_2", "TD549_3")
# 
# ssx <- vector(mode = "list", length = 9)
# dds <- vector(mode = "list", length = 9)
# res <- vector(mode = "list", length = 9)
# z <- vector(mode = "list", length = 9)
# vsd <- vector(mode = "list", length = 9)
# zz <- vector(mode = "list", length = 9)
# dgex <- vector(mode = "list", length = 9)
# sigx <- vector(mode = "list", length = 9)
# 
# for(i in 1:9){
#   print(paste("Processing comparison: ", i, " on 9"))
#   ssx[[i]] <- as.data.frame(colnames(yx[[i]]))
#   colnames(ssx[[i]]) <- "clone"
#   ssx[[i]]$DFT <- factor(c("DFT1","DFT1","DFT1","DFT2","DFT2","DFT2"))
#   dds[[i]] <- DESeqDataSetFromMatrix(countData = yx[[i]], colData = ssx[[i]],
#                                      design = ~ DFT)
#   res[[i]] <- DESeq(dds[[i]], quiet=TRUE)
#   z[[i]] <- results(res[[i]])
#   vsd[[i]] <- vst(dds[[i]])
#   zz[[i]] <- cbind(as.data.frame(z[[i]]), assay(vsd[[i]]))
#   dgex[[i]] <- as.data.frame(zz[[i]][order(zz[[i]]$pvalue),])
#   sigx[[i]] <- subset(dgex[[i]], padj < 0.05)
#   }
# 
# plotnames <- c("4906 vs RV", "4906 vs SN", "4906 vs TD549",
#                "1426 vs RV", "1426 vs SN", "1426 vs TD549",
#                "C5065 vs RV", "C5065 vs SN", "C5065 vs TD549")
# 
# for(i in 1:9){
#   maplot(dgex[[i]],plotnames[i])
#   make_volcano(dgex[[i]],plotnames[i])
#   
#   sigx[[i]][1:50,1:6] %>%
#   kbl(caption=plotnames[i]) %>%
#   kable_paper("hover", full_width = F)
#   
#   write.table(dgex[[i]],file=paste(plotnames[i],".tsv"),sep="\t")
#   
#   mx <- sigx[[i]][,7:ncol(sigx[[i]])]
#   mx <- head(mx,30)
#   colfunc <- colorRampPalette(c("blue", "white", "red"))
#   heatmap.2(as.matrix(mx),trace="none",scale="row",
#   col=colfunc(25),ColSideColors=cols[1:6],mar=c(8,14),
#   cexRow=0.7,cexCol=1)
# }

```

## Enrichment analysis

Need to get human homologs of these genes.
These were obtained from Ensembl v109 using Biomart.

```{r, hom}

rownames(dge1) <- sapply(strsplit(rownames(dge1),"\\."),"[[",1)
rownames(dge2) <- sapply(strsplit(rownames(dge2),"\\."),"[[",1)

# for(i in 1:9){
#   rownames(dgex[[i]]) <- sapply(strsplit(rownames(dgex[[i]]),"\\."),"[[",1)
# }


hm2 <- hm[hm$Tasmanian.devil.gene.stable.ID != "",]

gt <- hm2[,2:3]
length(unique(gt$Tasmanian.devil.gene.stable.ID))
length(unique(gt$Gene.name))

for(i in 1:length(gt$Gene.name)){
  if(gt$Gene.name[i]==""){gt$Gene.name[i] <- gt$Tasmanian.devil.gene.stable.ID[i]}
}

```

Now run mitch for DGE2. The pathways were sourced from Reactome on 7th July 2023.
```{r,mitch1, fig.dim = c(14, 10)}

genesets <- gmt_import("../ref/ReactomePathways_2023-07-14.gmt")

m2 <- mitch_import(dge2, DEtype="deseq2", geneTable=gt)
head(m2)

res2 <- mitch_calc(m2, genesets, priority="effect",cores=16)

if ( !file.exists("mitch_clines_reactome.html") ) {
  mitch_report(res2, "mitch_clines_reactome.html")
}

top <- subset(res2$enrichment_result,p.adjustANOVA<0.05)
saveRDS(top, "~/dftd_RNAseq_annelise/dge/gsea_clines.rds")

top %>%
  kbl(caption="Enriched pathways") %>%
  kable_paper("hover", full_width = F)

ggplot(top, aes(x = s.dist, y = reorder(set, s.dist), color = p.adjustANOVA, size = setSize)) + 
  geom_point(stat = 'identity') + 
  scale_color_viridis(option="mako", direction=1,limits = c(0, 0.05)) +
  xlab("s dist") + ylab("pathway") + ggtitle("Reactome ALL") + 
  theme_bw()

```

```{r,mitch2, fig.dim = c(8, 6)}

top20 <- rbind(slice_max(top, order_by=s.dist, n=10), slice_min(top, order_by=s.dist, n=10))

ggplot(top20, aes(x = s.dist, y = reorder(set, s.dist), color = p.adjustANOVA, size = setSize)) + 
  geom_point(stat = 'identity') + 
  scale_color_viridis(option="mako", direction=1,limits = c(0, 0.05)) +
  xlab("s dist") + ylab("pathway") + ggtitle("Reactome top 20") + 
  theme_bw()

```

```{r,mitch3, fig.dim = c(6, 4)}
react_metabosets <- readRDS("../ref/reactome_metabo.rds")

metabo1 <- filter(top, set %in% names(react_metabosets))
saveRDS(metabo1, "~/dftd_RNAseq_annelise/dge/gsea_clines_metabo.rds")

ggplot(metabo1 , aes(x = s.dist, y = reorder(set, s.dist), color = p.adjustANOVA, size = setSize)) + 
  geom_point(stat = 'identity') + 
  scale_color_viridis(option="mako", direction=1,limits = c(0, 0.05)) +
  xlab("s dist") + ylab("pathway") + ggtitle("Reactome metabolism") + 
  theme_bw()

top_reactome <- top

```

KEGG gene sets were sourced from MSigDb on 14 September 2023.

```{r,mitch4}

genesets2 <- gmt_import("../ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt.txt")

res2 <- mitch_calc(m2, genesets2, priority="effect",cores=16)

if ( !file.exists("mitch_clines_kegg.html") ) {
  mitch_report(res2, "mitch_clines_kegg.html")
}

top <- subset(res2$enrichment_result,p.adjustANOVA<0.05)

topup <- head(subset(top,s.dist>0),20)

topup %>%
  kbl(caption="Top 20 upregulated pathways") %>%
  kable_paper("hover", full_width = F)

topdn <- head(subset(top,s.dist<0),20)

topdn %>%
  kbl(caption="Top 20 downregulated pathways") %>%
  kable_paper("hover", full_width = F)

top2 <- rbind(head(topup,10),head(topdn,10))
top2 <- top2[order(-top2$s.dist),]
par(mar=c(5,15,5,2))

barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
  main="Top KEGG Pathways",cex.names=0.6,xlab="Enrichment Score",
  col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
  xlim=c(0,0.8))

ggplot(top, aes(x = s.dist, y = set, color = p.adjustANOVA, size = setSize)) + 
  geom_point(stat = 'identity') + 
  xlab("s dist") + ylab("pathway") + ggtitle("Reactome") + 
  theme_bw()

kegg_metabosets <- gmt_import("../ref/gsea/gsea_KEGG.gmt")

metabo2 <- filter(top, set %in% names(kegg_metabosets))
metabo2_ord <- metabo2[order(metabo2$s.dist),]
metabo2_ord$s.dist <- abs(metabo2_ord$s.dist)

barplot(metabo2_ord$s.dist,horiz=TRUE,las=1,names.arg=metabo2_ord$set,
  main="Enriched Metabolic Pathways - KEGG",cex.names=0.6,xlab="Enrichment Score",
  col=rev(c(rep("red",nrow(filter(metabo2, s.dist > 0))),rep("blue",nrow(filter(metabo2, s.dist < 0))))),
  xlim=c(0,0.8))


```
## Visualisation

Heatmaps.

```{r, complexheatmap1, fig.dim = c(4, 6)}

# see https://www.youtube.com/watch?v=ht1r34-ifVI for reference

# find corresponding human genes to devil ENS IDs
dge3 <- dge2 
dge3$ENSSHAG <- rownames(dge3)
hm3 <- data.frame(ENSSHAG=hm2[,2],geneID=hm2[,3])
dge3 <- left_join(dge3, hm3, by="ENSSHAG")
dge3 <- unite(dge3,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ", remove=F)
dge3 <- unique(dge3)
rownames(dge3) <- dge3$ENSSHAG_geneID

# only plot significant DEGs
dge3sig <- filter(dge3, padj < 0.05)
dge3sig <- dge3sig[order(dge3sig$log2FoldChange, decreasing=TRUE),]

mx <- dge3sig[,7:ncol(dge3sig)] # get normalised counts
 
mx.scaled <- t(apply(mx[,1:6], 1, scale)) # center and scale each column (Z-score)
colnames(mx.scaled) <- c("DFT1 4906", "DFT1 C5065", "DFT1 1426", "DFT2 RV", "DFT2 SN", "DFT2 TD549")
log2FC <- as.matrix(dge3sig$log2FoldChange)
rownames(log2FC) <- rownames(dge3sig)
colnames(log2FC) <- "logFC"
mean <- as.matrix(dge3sig$baseMean)
rownames(mean) <- rownames(dge3sig)
colnames(mean) <- "AveExpr"

col_logFC <- colorRamp2(c(min(log2FC),0,max(log2FC)), hcl_palette = "Blue-Red 2")
#col_AveExpr <- colorRamp2(c(quantile(mean)[1], quantile(mean)[4]), c("white","red"))

ha <- HeatmapAnnotation(summary=anno_summary(gp=gpar(fill="#D5D7E2"),height=unit(1.5, "cm")))

# ----- Top DEGs in REACTOME metabolic pathways ------

metabo <- c() # get all names of genes involved in metabolism
for(i in 1:length(genesets)){metabo <- c(metabo,genesets[[i]])}
metabo <- unique(metabo) # get rid of duplicates

temp <- c() # get ENSSHAG and gene ID
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

# top 20 up & down
met40 <- rbind(head(met,20),tail(met,20))
met40FC <- as.matrix(c(head(metFC,20),tail(metFC,20)))
rownames(met40FC) <- rownames(met40)
colnames(met40FC) <- "logFC"
met40AE <- as.matrix(c(head(metAE,20),tail(metAE,20)))
rownames(met40AE) <- rownames(met40)
colnames(met40AE) <- "AveExpr"

custom_heatmap(met40, met40FC, met40AE, title = "TOP 40 - REACTOME metabolism")

# ----- All DEGs in REACTOME metabolic pathways ------

dge3_metabo_reactome <- filter(dge3, geneID %in% metabo)

make_volcano(dge3_metabo_reactome, "Cell lines - Metabolism genes - REACTOME")

# ----- Individual TOP REACTOME metabolic pathways -----

pathways <- metabo1$set
heatmaps <- vector(mode = "list", length = length(pathways))
for(i in 1:length(pathways)){
  
  metabo <- c() # get all names of genes involved in metabolism
  for(j in 1:length(genesets)){metabo <- c(metabo,genesets[[pathways[i]]])}
  metabo <- unique(metabo) # get rid of duplicates

  temp <- c() # get ENSSHAG and gene ID
  for(k in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[k]))}
  metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

  met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
  metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
  metAE <- subset(mean, rownames(mean) %in% metabolism)

  # react_pathw[[i]] <- custom_heatmap(met, metFC, metAE, title = paste(pathways[i], "- REACTOME"))
  heatmaps[[i]] <- custom_heatmap(met, metFC, metAE, title = paste(pathways[i]))
  
}

```

```{r,complexheatmap_1, fig.dim = c(4, 3.25)}
print(heatmaps[[1]])
```

```{r,complexheatmap_2, fig.dim = c(4, 3.75)}
print(heatmaps[[2]])
```

```{r,complexheatmap_3, fig.dim = c(4, 4)}
print(heatmaps[[3]])
```

```{r,complexheatmap_4, fig.dim = c(4, 5.5)}
print(heatmaps[[4]])
```

```{r, other, fig.dim = c(4, 5.75)}
# ----- Other interesting pathways REACTOME -----

mm <- filter(hm3, geneID %in% genesets[["Glycolysis"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Glycolysis")
```

```{r, other4, fig.dim = c(4, 5.75)}
mm <- filter(hm3, geneID %in% genesets[["Respiratory electron transport, ATP synthesis by chemiosmotic coupling, and heat production by uncoupling proteins."]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Respiratory electron transport, \n ATP synthesis by chemiosmotic coupling, \n and heat production by uncoupling proteins.")
```
```{r, other5, fig.dim = c(4, 3.25)}

mm <- filter(hm3, geneID %in% genesets[["Citric acid cycle (TCA cycle)"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Citric acid cycle (TCA cycle)")
```
```{r, other6, fig.dim = c(4, 5.5)}
mm <- filter(hm3, geneID %in% genesets[["Respiratory electron transport"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Respiratory electron transport")
```
```{r, other7, fig.dim = c(4, 4)}
mm <- filter(hm3, geneID %in% genesets[["DNA Repair"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

#custom_heatmap(met, metFC, metAE, title = "DNA Repair")
```

```{r, other8, fig.dim = c(4, 6.5)}
mm <- filter(hm3, geneID %in% genesets[["DNA Double-Strand Break Repair"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

#custom_heatmap(met, metFC, metAE, title = "DNA Double-Strand Break Repair")
```

```{r, other9, fig.dim = c(4, 4)}
mm <- filter(hm3, geneID %in% genesets[["Detoxification of Reactive Oxygen Species"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Detoxification of Reactive Oxygen Species")
```
```{r, other10, fig.dim = c(4, 3)}
mm <- filter(hm3, geneID %in% genesets[["Pentose phosphate pathway"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Pentose phosphate pathway")
```

```{r, complexheatmap2, fig.dim = c(4, 8.5)}

mm <- filter(hm3, geneID %in% genesets[["Fatty acid metabolism"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Fatty acid metabolism")

```

```{r, complexheatmap5,fig.dim = c(4, 2.75)}
mm <- filter(hm3, geneID %in% genesets[["Glutamate and glutamine metabolism"]])
mm <- c(t(unite(mm,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% mm)
metFC <- subset(log2FC, rownames(log2FC) %in% mm)
metAE <- subset(mean, rownames(mean) %in% mm)

custom_heatmap(met, metFC, metAE, title = "Glutamate and glutamine metabolism")
```

```{r, complexheatmap6,fig.dim = c(4, 6)}
# ----- Top DEGs in KEGG metabolic pathways ------

metabo <- c()
for(i in 1:length(genesets2)){metabo <- c(metabo,genesets2[[i]])}
metabo <- unique(metabo)

temp <- c()
for(i in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[i]))}
metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
metAE <- subset(mean, rownames(mean) %in% metabolism)

# top 20 up & down
met40 <- rbind(head(met,20),tail(met,20))
met40FC <- as.matrix(c(head(metFC,20),tail(metFC,20)))
rownames(met40FC) <- rownames(met40)
colnames(met40FC) <- "logFC"
met40AE <- as.matrix(c(head(metAE,20),tail(metAE,20)))
rownames(met40AE) <- rownames(met40)
colnames(met40AE) <- "AveExpr"

custom_heatmap(met40, met40FC, met40AE, title = "TOP 40 - KEGG metabolism")

# ----- All DEGs in KEGG metabolic pathways ------

dge3_metabo_kegg <- filter(dge3, geneID %in% metabo)

make_volcano(dge3_metabo_kegg, "Cell lines - Metabolism genes - KEGG")

# ----- Individual TOP KEGG metabolic pathways -----

pathways <- metabo2$set

for(i in 1:length(pathways)){
  
  metabo <- c() # get all names of genes involved in metabolism
  for(j in 1:length(genesets2)){metabo <- c(metabo,genesets2[[pathways[i]]])}
  metabo <- unique(metabo) # get rid of duplicates

  temp <- c() # get ENSSHAG and gene ID
  for(k in 1:length(metabo)){temp <- rbind(temp,filter(hm3, geneID==metabo[k]))}
  metabolism <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

  met <- subset(mx.scaled, rownames(mx.scaled) %in% metabolism)
  metFC <- subset(log2FC, rownames(log2FC) %in% metabolism)
  metAE <- subset(mean, rownames(mean) %in% metabolism)

  # react_pathw[[i]] <- custom_heatmap(met, metFC, metAE, title = paste(pathways[i], "- REACTOME"))
  print(custom_heatmap(met, metFC, metAE, title = paste(pathways[i], "- KEGG")))
  
}

# Sebastian's hypothesis genes
# GDNF, NGF, NRTN, ARTN, GFRA1, NTF3, CCL2, CX3CL1

seb_genes <- filter(dge3, geneID %in% c("GDNF", "NGF", "NRTN", "ARTN", "GFRA1", "NTF3", "CCL2", "CX3CL1"))

sig <- subset(seb_genes,padj<0.05)

# heatmap
mx <- sig[,7:12]
mx <- head(mx,30)
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2(as.matrix(mx),trace="none",scale="row",
  col=colfunc(25),mar=c(5,14),
  cexRow=0.8,cexCol=0.7)

```

#### Term-gene graph

```{r, tgg}

degenes <- data.frame(FC=dge3sig$log2FoldChange, ID=dge3sig$geneID)

paths <- c("Glycolysis",
           "Formation of ATP by chemiosmotic coupling",
           "Citric acid cycle (TCA cycle)",
           "Respiratory electron transport",
           "DNA Double-Strand Break Repair",
           "Detoxification of Reactive Oxygen Species",
           "Pentose phosphate pathway",
           "Fatty acid metabolism",
           "Glutamate and glutamine metabolism")

path_list <- list()

for(i in 1:nrow(metabo1)){
  path_temp <- metabo1$set[i]
  genes_temp <- genesets[[path_temp]]
  list_temp <- list(intersect(genes_temp,degenes$ID))
  path_list <- append(path_list, list_temp)
}

names(path_list) <- metabo1$set

degenes <- na.omit(data.frame(ID=dge3sig$geneID,FC=dge3sig$log2FoldChange))

mydf_2 <- degenes %>% group_by(ID) %>% summarize(FC=mean(FC))
mydf_2 <- data.frame(mydf_2)
mydf_2 <- mydf_2[-1,]
mydf_3 <- mydf_2[,-1]

FC <- setNames(mydf_3, mydf_2$ID)

cnetplot(path_list, foldChange=FC) 



path_list <- list()

for(i in 1:length(paths)){
  path_temp <- paths[i]
  genes_temp <- genesets[[path_temp]]
  list_temp <- list(intersect(genes_temp,degenes$ID))
  path_list <- append(path_list, list_temp)
}

names(path_list) <- paths

degenes <- na.omit(data.frame(ID=dge3sig$geneID,FC=dge3sig$log2FoldChange))

mydf_2 <- degenes %>% group_by(ID) %>% summarize(FC=mean(FC))
mydf_2 <- data.frame(mydf_2)
mydf_2 <- mydf_2[-1,]
mydf_3 <- mydf_2[,-1]

FC <- setNames(mydf_3, mydf_2$ID)

# TODO: show TCA+ETC and FOA+ROS
cnetplot(path_list, foldChange=FC, showCategory = 3) 

```


```{r, visualise pathways}
#setwd("/mnt/data/annelise/dftd_RNAseq_annelise/dge/pathways_visu")
data("sbgn.xmls")
data(pathways.info)
input.pathways <- data.frame(findPathways(c(names(path_list),metabo1$set)))
SBGNview.obj <- SBGNview(
          gene.data = FC, 
          gene.id.type = "SYMBOL",
          input.sbgn = input.pathways$pathway.id,
          output.file = "clines", 
          output.formats =  c("svg"),
          min.gene.value = -10,
          max.gene.value = 10)
print(SBGNview.obj)
```

Run mitch on cell line pairwise comparisons.

```{r, mitchx}

# mx <- vector(mode = "list", length = 9)
# resx <- vector(mode = "list", length = 9)
# 
# for(i in 1:9){
#   
#   mx[[i]] <- mitch_import(dgex[[i]], DEtype="deseq2",geneTable=gt)
#   resx[[i]] <- mitch_calc(mx[[i]], genesets3, priority="effect",cores=16)
# 
#   if ( !file.exists(paste(plotnames[i],"_mitch.html")) ) {
#     mitch_report(resx[[i]], paste(plotnames[i],"_mitch.html"))
#   }
#   
#   top <- subset(resx[[i]]$enrichment_result, p.adjustANOVA<0.05)
#   
#   topup <- head(subset(top,s.dist>0),20)
#   
#   topup %>%
#   kbl(caption=paste("Top 20 upregulated metabolic pathways: ", plotnames[i])) %>%
#   kable_paper("hover", full_width = F)
#   
#   topdn <- head(subset(top,s.dist<0),20)
#   
#   topdn %>%
#     kbl(caption=paste("Top 20 downregulated metabolic pathways: ", plotnames[i])) %>%
#     kable_paper("hover", full_width = F)
#   
#   top2 <- rbind(head(topup,10),head(topdn,10))
#   top2 <- top2[order(-top2$s.dist),]
#   par(mar=c(5,15,5,2))
#   
#   barplot(rev(abs(top2$s.dist)),horiz=TRUE,las=1,names.arg=rev(top2$set),
#           main=paste(plotnames[i]),
#           cex.names=0.6,xlab="Enrichment Score",
#           col=c(rep("blue",nrow(head(topdn,10))),rep("red",nrow(head(topup,10)))),
#           xlim=c(0,0.8))
#   
# }

```

Make heatmaps for pairwise cell line comparisons.

```{r, complexheatmap4}

# ----- Fatty acid metabolism heatmaps -----
# fatty1 <- genesets2[["REACTOME_FATTY_ACID_METABOLISM"]]
# fatty2 <- genesets2[["HALLMARK_FATTY_ACID_METABOLISM"]]
# fatty3 <- genesets2[["GOBP_FATTY_ACID_BIOSYNTHETIC_PROCESS"]]
# fatty4 <-genesets2[["KEGG_FATTY_ACID_METABOLISM"]]
# fatty <- list(fatty1=fatty1,fatty2=fatty2,fatty3=fatty3,fatty4=fatty4)
# title <- c("REACTOME_FATTY_ACID_METABOLISM","HALLMARK_FATTY_ACID_METABOLISM","GOBP_FATTY_ACID_BIOSYNTHETIC_PROCESS","KEGG_FATTY_ACID_METABOLISM")
# 
# for(i in 1:length(fatty)){
#   temp <- c()
#   for(j in 1:length(fatty[[i]])){temp <- rbind(temp,filter(hm3, geneID==fatty[[i]][j]))}
#   fattyx <- c(t(unite(temp,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))
#   fat <- subset(mx.scaled, rownames(mx.scaled) %in% fattyx)
#   fatFC <- subset(log2FC, rownames(log2FC) %in% fattyx)
#   fatAE <- subset(mean, rownames(mean) %in% fattyx)
#   
#   h1 <- Heatmap(fat, cluster_rows = F,
#                 column_labels = colnames(fat), name="Z-score",
#                 cluster_columns = T)
#   h2 <- Heatmap(fatFC, row_labels = rownames(fatFC), row_names_gp = gpar(fontsize = 6),
#               cluster_rows = F, name="logFC", top_annotation = ha, col = col_logFC,
#               cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
#                 grid.text(round(as.numeric(fatFC[i, j],2)), x, y, gp=gpar(fontsize=6))
#               })
#   h3 <- Heatmap(fatAE, row_labels = rownames(fatAE), row_names_gp = gpar(fontsize = 6),
#               cluster_rows = F, name = "AveExpr", col=col_AveExpr,
#               cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
#                 grid.text(round(as.numeric(fatAE[i, j],2)), x, y, gp=gpar(fontsize=6))},
#               column_title=title[i])
#   
#   name <- paste("q",i,sep="")
#   assign(name, h1 + h2 + h3)
# }
# 
# q1
# q2
# q3
# q4

```

```{r, Venn data}

allgenes <- rownames(dge)
sigdegs <- rownames(sig2)
updegs <- sig2_up
dndegs <- sig2_dn

metreactgenes <- rownames(dge3_metabo_reactome)
metsigreactdegs <- rownames(filter(dge3_metabo_reactome, padj < 0.05))
metsigreactupdegs <- rownames(filter(filter(dge3_metabo_reactome, padj < 0.05),log2FoldChange > 0))
metsigreactdndegs <- rownames(filter(filter(dge3_metabo_reactome, padj < 0.05),log2FoldChange < 0))
metkegggenes <- rownames(dge3_metabo_kegg)
metsigkeggdegs <- rownames(filter(dge3_metabo_kegg, padj < 0.05))
metsigkeggupdegs <- rownames(filter(filter(dge3_metabo_kegg, padj < 0.05),log2FoldChange > 0))
metsigkeggdndegs <- rownames(filter(filter(dge3_metabo_kegg, padj < 0.05),log2FoldChange < 0))

venn_clines <- list(allgenes=allgenes, sigdegs=sigdegs, updegs=updegs, dndegs=dndegs, metreactgenes=metreactgenes,
                      metsigreactdegs=metsigreactdegs, metsigreactupdegs=metsigreactupdegs, metsigreactdndegs=metsigreactdndegs,
                      metkegggenes=metkegggenes, metsigkeggdegs=metsigkeggdegs, metsigkeggupdegs=metsigkeggupdegs,
                      metsigkeggdndegs=metsigkeggdndegs)
saveRDS(venn_clines, file = "venn_clines.rds")

saveRDS(dge2, "~/dftd_RNAseq_annelise/dge/dge_clines.rds")

# venn.diagram(
#   x = list(alldegs, updegs, downdegs),
#   category.names = c("All DEGs" , "Up", "Down"),
#   filename = 'updown_venn_diagramm.png',
#   output=TRUE
# )

```

## Adding custom gene sets

```{r, custom_g, fig.dim = c(4, 3.25)}

cell_compet_genes <- read_excel("../ref/gsea/Cell_compet_genes.xlsx")
names(cell_compet_genes)[names(cell_compet_genes) == "Ensembl_ID"] <- "Gene.stable.ID"

cell_compet_genes <- inner_join(cell_compet_genes,hm,by="Gene.stable.ID") # making sure these genes exist in devils

genesets$"Cell Competition" <- unique(cell_compet_genes$Gene.name)
genesets$"Cell Competition - Mechanical Competition" <- unique(filter(cell_compet_genes, Type == "Mechanical_Competition")$Gene.name)

metastasis_genes <- read_excel("../ref/gsea/Metastasis_genes.xlsx")
names(metastasis_genes)[names(metastasis_genes) == "Ensembl_ID"] <- "Gene.stable.ID"

metastasis_genes <- inner_join(metastasis_genes,hm,by="Gene.stable.ID") # making sure these genes exist in devils

genesets$"Metastasis" <- unique(metastasis_genes$Gene.name)
genesets$"Metastasis - promotors" <- unique(filter(metastasis_genes, Type == "Metastasis_promotor")$Gene.name)
genesets$"Metastasis - suppressors" <- unique(filter(metastasis_genes, Type == "Metastasis_supressor")$Gene.name)

# gsea
res_custom <- mitch_calc(m2, genesets, priority="effect",cores=16)

if ( !file.exists("mitch_custom.html") ) {
  mitch_report(res_custom, "mitch_custom.html")
}


cc <- filter(hm3, geneID %in% genesets[["Cell Competition"]])
cc <- c(t(unite(cc,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% cc)
metFC <- subset(log2FC, rownames(log2FC) %in% cc)
metAE <- subset(mean, rownames(mean) %in% cc)

custom_heatmap(met, metFC, metAE, title = "Cell competition")
```
```{r, custom_g2, fig.dim = c(4, 5)}
meta <- filter(hm3, geneID %in% genesets[["Metastasis"]])
meta <- c(t(unite(meta,"ENSSHAG_geneID", ENSSHAG:geneID, sep=" ")))

met <- subset(mx.scaled, rownames(mx.scaled) %in% meta)
metFC <- subset(log2FC, rownames(log2FC) %in% meta)
metAE <- subset(mean, rownames(mean) %in% meta)

custom_heatmap(met, metFC, metAE, title = "Metastasis")

```

## Session information

For reproducibility.

```{r,session}

sessionInfo()

```
