---
title: "Cell line vs biopsies"
author: "Anne-Lise Gérard"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 3
    fig_height: 3
theme: cosmo
---

Source codes: https://github.com/Anne-LiseGerard/dftd_rnaseq

```{r,pkg}

suppressPackageStartupMessages({
  library("VennDiagram")
  library("ggplot2")
  library("mitch")
  library("dplyr")
})

```

## Background

Goal: compare expression profiles of DFT1 vs DFT2 tumours and cell lines.

## Functions

```{r, func}

display_venn <- function(x, ...){
  library(VennDiagram)
  grid.newpage()
  venn_object <- venn.diagram(x, filename = NULL, ...,
                              height = 480 , width = 480 , 
                              resolution = 300,
                              lwd = 2,
                              col=c('darkseagreen', 'darksalmon'),
                              fill = c(alpha('darkseagreen',0.3), alpha('darksalmon',0.3)),
                              cex = 1,
                              fontfamily = "sans",
                              cat.cex = 0.9,
                              cat.default.pos = "outer",
                              cat.dist = c(0.05, 0.05),
                              cat.pos = c(-25,25),
                              cat.fontfamily = "sans",
                              cat.col = c('darkseagreen', 'darksalmon'),
                              main.cex = 1,
                              main.fontfamily = "sans",
                              sub.cex = 0.8,
                              sub.fontfamily = "sans",
                              ext.text=FALSE)
  grid.draw(venn_object)
}

# get rid of log files
futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")

```

## Load data

Import lists of genes for Venn diagrams , DESEq2 outputs and genesets. 

```{r, import}

clines <- readRDS("~/dftd_RNAseq_annelise/dge/venn_clines.rds") # list genes cell lines
biopsies <- readRDS("~/dftd_RNAseq_annelise/dge/venn_biopsies.rds") # list genes biopsies

dge_clines <- readRDS("~/dftd_RNAseq_annelise/dge/dge_clines.rds") # DESeq2 cell lines
dge_biopsies <- readRDS("~/dftd_RNAseq_annelise/dge/dge_biopsies.rds") # DESeq2 biopsies

gt <- readRDS("~/dftd_RNAseq_annelise/dge/gt.rds") # homology
genesets <- gmt_import("~/dftd_RNAseq_annelise/ref/ReactomePathways_2023-07-14.gmt") # reactome
genesets2 <- gmt_import("~/dftd_RNAseq_annelise/ref/c2.cp.kegg.v2023.1.Hs.symbols.gmt.txt") # kegg

```

## Visualisation

```{r, graphs}

display_venn(x=list(clines[["allgenes"]], biopsies[["allgenes"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="All genes")
display_venn(x=list(clines[["sigdegs"]], biopsies[["sigdegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="DEGs",
             sub="(padj < 0.05)")
display_venn(x=list(clines[["updegs"]], biopsies[["updegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Up DEGs",
             sub="(padj < 0.05; logFC > 0)")
display_venn(x=list(clines[["dndegs"]], biopsies[["dndegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Down DEGs",
             sub="(padj < 0.05; logFC < 0)")

# genes related to metabolism
display_venn(x=list(clines[["metreactgenes"]], biopsies[["metreactgenes"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism genes - REACTOME")
display_venn(x=list(clines[["metsigreactdegs"]], biopsies[["metsigreactdegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism DEGs - REACTOME",
             sub="(padj < 0.05)")
display_venn(x=list(clines[["metsigreactupdegs"]], biopsies[["metsigreactupdegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism up DEGs - REACTOME",
             sub="(padj < 0.05; logFC > 0)")
display_venn(x=list(clines[["metsigreactdndegs"]], biopsies[["metsigreactdndegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism down DEGs - REACTOME",
             sub="(padj < 0.05; logFC < 0)")

display_venn(x=list(clines[["metkegggenes"]], biopsies[["metkegggenes"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism genes - KEGG")
display_venn(x=list(clines[["metsigkeggdegs"]], biopsies[["metsigkeggdegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism DEGs - KEGG",
             sub="(padj < 0.05)")
display_venn(x=list(clines[["metsigkeggupdegs"]], biopsies[["metsigkeggupdegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism up DEGs - KEGG",
             sub="(padj < 0.05; logFC > 0)")
display_venn(x=list(clines[["metsigkeggdndegs"]], biopsies[["metsigkeggdndegs"]]),
             category.names = c("Cell lines" , "Biopsies"),
             main="Metabolism down DEGs - KEGG",
             sub="(padj < 0.05; logFC < 0)")


```

## Enrichment

The pathway 'Response to metal ions' causes a bug as the genenames appear to map to the same
human gene.
I think the duplicate data is causing problems.

```{r, mitch}

x <- list("de_clines"=dge_clines, "de_biopsies"=dge_biopsies)

y <- mitch_import(x, "deseq2", geneTable = gt)

# REACTOME

setnames <-c( "Response to metal ions", "Metallothioneins bind metals")
genesets <- genesets[! names(genesets) %in% setnames]

res_react <- mitch_calc(y, genesets)

if ( !file.exists("mitch_comp_react.html") ) {
  mitch_report(res_react, "mitch_comp_react.html")
}

# KEGG
res_kegg <- mitch_calc(y, genesets2)

if ( !file.exists("mitch_comp_kegg.html") ) {
  mitch_report(res_kegg, "mitch_comp_kegg.html")
}

```


## Session information

```{r,session}

sessionInfo()

```
